---
output: pdf_document
---

# Relationships

In our final level inference we correlate values across time rather than correlating values from a single moment or a single averaged moment. For example, we might ask if affect is related to performance across time; i.e., when affect is high is performance also high, and when affect is low is performance also low? 

This inference sounds similar to the one just presented, but their difference is important. With inference three we ask about affect and performance at $t$ or at an averaged window of $t$ -- we examine, for example, how ending performance relates to ending affect, or how affect averaged across time relates to performance averaged across time. Here, we retain all of the information and examine the relationship between affect and performance across all $t$. 

Figure \ref{level_relation} shows this inference graphically. In Panel A we plot affect and performance trajectories across time, where the solid line indicates affect and the dashed line indicates performance -- this time we only focus on a single individual or unit. The colored squares represent levels at different points in time. The green squares highlight low values of both variables, the blue high values, and the red middle values. 

Panel B shows how those respective values map onto a scatterplot of affect and performance -- which again will lead to the inference. The blue values indicate that high values of affect tend to co-occur with high values of performance (shown respectively by the blue squares in Panel A). The red values indicate that middle values of affect tend to co-occur with middle values of performance. The green values, finally, indicate that low values of affect tend to co-occur with low values of performance. Across time, affect and performance covary. 


Inference four, finally, is about the relationship between raw values across time.

>> **Inference 1:** There is a relationship between $x$ and $y$ across time. 

Figure \ref{relation_tvc} shows the inference.

>> **Inference 2:** There is a relationship between $x$ and $y$ across units over time.

## Relationships Inference Table

Text.

\begin{center}

------------

Insert Table \ref{relationships_table} about here

------------

\end{center}

## Models

Time-varying or invariant covariates analyses.




```{r, echo = F}
library(xtable)
library(knitr)
library(kableExtra)
library(tidyverse)

inference_text <- c('There is a relationship between x and y across time.',
                    'There is a relationship between x and y across units over time')
level_table <- data.frame(
  Inference = c(1, 2),
  Examples = c(linebreak('Affect relates to performance across time.\nHelping behaviors predict depletion across time.'),
                         'Affect relates to performance across people (and it is consistent over time/and averaged over time it is positive).'))

level_table %>%
  kable('latex', booktabs = T, escape = F, align = 'l', caption = '\\label{relationships_table}Examples of Relationship inferences.') %>%
    column_spec(1, width = '5em') %>%
    column_spec(2, width = '30em') %>%
  row_spec(1, hline_after = T)


```


```{r, echo = F, fig.cap = 'Relating affect to performance on one unit across time.\\label{level_relation}', fig.height = 8}
library(tidyverse)
set.seed(2)
timer <- 20
da <- matrix(, ncol = 3, nrow = timer)

for(i in 1:timer){
  
  if(i < 6){
  da[i, 1] <- i
  da[i, 2] <- 3 + rnorm(1, 0, 4)
  da[i, 3] <- 19 + rnorm(1, 0, 1) + 0.6* da[i, 2]
  }else if (i > 5 & i < 11){
    
    da[i, 1] <- i
    da[i, 2] <- 16 + rnorm(1,0,1)
    da[i, 3] <- 32 + rnorm(1,0,1) + 0.6*da[i, 2]
  } else{
    
      da[i, 1] <- i
      da[i, 2] <- 3 + rnorm(1, 0, 4)
      da[i, 3] <- 6 + rnorm(1, 0, 1) + 0.6* da[i, 2]
    
  }
}

da <- data.frame(da)
names(da) <- c('Time', 'Affect', 'Performance')
da <- da %>%
  gather(Affect, Performance, key = 'Variable', value = 'Value')

library(ggplot2)

g1 <- ggplot(da, aes(x = Time, y = Value, group = Variable)) + 
  geom_line(aes(linetype = Variable)) + 
  theme_classic() + 
  scale_y_continuous(labels = NULL) + 
  geom_rect(aes(xmin = 2, xmax = 4, ymin = 3, ymax = 6), fill = 'red', alpha = 0.03) + 
  #geom_segment(aes(x = 3, xend = 3, y = 5, yend = 21), color = 'blue', alpha = 0.5, linetype = 'dashed') + 
  geom_rect(aes(xmin = 2, xmax = 4, ymin = 20, ymax = 23), fill = 'red', alpha = 0.03) + 
  geom_rect(aes(xmin = 18, xmax = 20, ymin = 3, ymax = 6), fill = 'green', alpha = 0.03) + 
  #geom_segment(aes(x = 17, xend = 17, y = 19, yend = 32), color = 'blue', alpha = 0.5, linetype = 'dashed') + 
  geom_rect(aes(xmin = 18, xmax = 20, ymin = -2, ymax = 1), fill = 'green', alpha = 0.03) + 
  geom_rect(aes(xmin = 7, xmax = 9, ymin = 40, ymax = 43), fill = 'blue', alpha = 0.03) + 
  geom_rect(aes(xmin = 7, xmax = 9, ymin = 16, ymax = 19), fill = 'blue', alpha = 0.03) + 
  ggtitle('Panel A')

affect_line <- c(1:21)
perf_line <- affect_line*0.7 + rnorm(21, 0, 1)
lin_df <- data.frame(
  'Affect' = c(affect_line),
  'Performance' = c(perf_line),
  'label' = c(rep('green', 7), rep('red', 7), rep('blue', 7)))

g2 <- ggplot(lin_df, aes(x = Affect, y = Performance, color = label)) + 
  geom_point(size = 2, alpha = 0.6) + 
  theme_classic() + 
  scale_x_continuous(labels = NULL) + 
  scale_y_continuous(labels = NULL) + 
  scale_color_manual(values = c('blue' = 'blue',
                                'green' = 'green',
                                'red' = 'red')) + 
  theme(legend.position = 'none') + 
  ggtitle('Panel B')

library(gridExtra)

grid.arrange(g1, g2, ncol = 1)
```




```{r, echo = F, fig.cap = 'Relating affect to performance across units over time. A typical time-varying covariates model constrains the correlation to be equivalent across time. Here, the relationship is unconstrained at each time point.\\label{relation_tvc}', fig.height = 8}
library(ggplot2)
set.seed(94)

time <- c(1:20)
r_affect <- c(8 + rnorm(20,0,1))
r_perf <- c(6 + rnorm(20,0,1))

r_affect2 <- c(2 + rnorm(20,0,1))
r_perf2 <- c(4 + rnorm(20,0,1))

r_affect3 <- c(5 + rnorm(20,0,1))
r_perf3 <- c(4.8 + rnorm(20,0,1))

rel_df <- data.frame(
  'Time' = c(time, time, time, time, time, time),
  'Value' = c(r_affect, r_perf,
          r_affect2, r_perf2,
          r_affect3, r_perf3),
  'call' = c(rep('Affect', 20),
              rep('Performance', 20),
             rep('Affect', 20),
             rep('Performance', 20),
             rep('Affect', 20),
             rep('Performance', 20)),
  'Person' = c(rep('Person 1', 40),
               rep('Person 2', 40),
               rep('Person 3', 40)))


library(tidyverse)

untidy_df <- rel_df %>% 
    spread(Value, key = call) 

untidy_df <- untidy_df %>%
  group_by(Time) %>%
  summarise(
    cor_ap = cor(Affect, Performance)
  )


library(data.table)
selector <- data.table(rel_df)
#selector[selector[, Person == 'Person 1' & Time == 6 & call == 'Affect'], 2]
#as.numeric(selector[selector[, Person == 'Person 1' & Time == 6 & call == 'Affect'], 2])



label_data <- data.frame(
  'Time' = c(6,6,6,6,6,6),
  'Label' = c(
    # affect person 1
    as.numeric(selector[selector[, Person == 'Person 1' & Time == 6 & call == 'Affect'], 2]),
    # affect person 2
    as.numeric(selector[selector[, Person == 'Person 2' & Time == 6 & call == 'Affect'], 2]),
    # affect person 3
    as.numeric(selector[selector[, Person == 'Person 3' & Time == 6 & call == 'Affect'], 2]),
    # performance person 1
    as.numeric(selector[selector[, Person == 'Person 1' & Time == 6 & call == 'Performance'], 2]),
    # performance person 2
    as.numeric(selector[selector[, Person == 'Person 2' & Time == 6 & call == 'Performance'], 2]),
    # performance person 3
    as.numeric(selector[selector[, Person == 'Person 3' & Time == 6 & call == 'Performance'], 2])

  ),
  'call' = c('Affect', 'Affect','Affect',
             'Performance', 'Performance', 'Performance'),
  'Person' = c('Person 1', 'Person 2', 'Person 3', 'Person 1', 'Person 2', 'Person 3')
  
)

relate1 <- ggplot(rel_df, aes(x = Time, y = Value, group = call)) + 
  #geom_smooth(aes(linetype = call), se = F, color = 'black', span = 0.4) + 
  geom_line(aes(linetype = call)) + 
  theme_classic() + 
  scale_y_continuous(labels = NULL) + 
  theme(legend.title = element_blank()) + 
  facet_grid(~Person, scales = 'free', space = 'free') + 
  geom_point(data = label_data, size = 4, colour = 'red', aes(y = Label), show.legend = F, alpha = 0.6) + 
  ggtitle('Panel A')


point_it <- data.frame(
  'Time' = c(6),
  'boom' = c(untidy_df[untidy_df$Time == 6, 2])
)
names(point_it) <- c('Time', 'boom')

relate2 <- ggplot(untidy_df, aes(x = Time, y = cor_ap)) + 
  geom_point(size = 4, alpha = 0.6) + 
  theme_classic() + 
  scale_y_continuous(breaks = c(0, 1)) + 
  ylab("Affect and Performance Correlation Across People") + 
  xlab("Time") + 
  ggtitle('Panel B') + 
  theme(legend.title = element_blank()) + 
  theme(
    axis.text.y = element_blank(),
    axis.ticks = element_blank()
  ) + 
  geom_point(data = point_it, color = 'red', size = 4, aes(x = Time, y = boom), show.legend = F, alpha = 0.6)

library(gridExtra)

grid.arrange(relate1, relate2, ncol = 1)

```