---
output: pdf_document
---

# Trend

Does affect go up or down across time, or is it relatively stable? Does trainee skill increase over the training session? These are questions about trend, and these first two are specifically about linear trend. It is also possible to explore how variables bend or curve across time. Do newcomer perceptions of climate increase and then plateau over time? Does the response time of a medical team decrease with each successive case but then remain stable once the team can no longer improve their coordination? These latter questions concern curvilinear trajectories. 

Trend has to do with the global shape of the trajectory across time. If we put a variable on the $y$-axis and plot its values against time on the $x$-axis, do the values tend to go up or down over time? It can be thought of as the coarse-grained direction of a trajectory. 

Figure \ref{trend} demonstrates trend, where the red line shows positive, increasing trend, the blue line shows negative, decreasing trend, and the green line shows a curvilinear trajectory (ADD GREEN LINE). Keep in mind that curvilinear and linear trajectories are both *linear in parameters* and should not be confused with non-linear systems. 

\begin{center}

------------

Insert Figure \ref{trend} about here

------------

\end{center}

Our first trend inference, therefore, concerns the shape of the trajectory.

>> **Inference 1:** There is positive/negative/curvilinear trend in a variable across time. 

As with the level inferences, when we add more units we can examine trend variability. Do all trainees develop greater skill across time? Is there variability in the trend of helping behaviors, or counterproductive work behaviors over time? 

Figure \ref{trend_var} shows differences in trend variability. In the first panel all units (people) show the same positive trend, whereas everyone in the second panel shows different trend: person one's data appear to increase over time, person two's data remain flat, and person three's data decrease over time. With greater variability there is less consistency in trend across units.

\begin{center}

------------

Insert Figure \ref{trend_var} about here

------------

\end{center}

>> **Inference 2:** There is variability in the trend of a variable across time; trend differs across units.

Inferences one and two are about one variable, but they can also be iterated across all observed variables. For example, we might discover that affect and performance trends both decrease, but there is greater variability across units in the affect trend. Or we might learn that affect has a negative trend while performance has a positive trend.

Correlating these trends is the next inference. Correlating indicates co-occuring patterns, but this time we are focused on trends rather than levels. A large positive correlation between affect and performance trends indicates that people with a positive affect trend (usually) have a positive performance trend and people with a negative affect trend (usually) have a negative performance trend. 

Figure \ref{trend_correlation} shows the inuition behind the inference with a set of graphs. On the top panel, affect and performance are plotted across time for three individuals. For individual one, affect goes up while performance goes down, this pattern is reversed for person two, and person three reports fluctuating affect and performance but with no trend. The bottom panel then maps those pairings onto a figure that shows the relationship between the affect and performance trend. For example, person one has a positive affect trend and a negative performance trend, so their point on the bottom panel goes on the bottom right, whereas person two has the opposite pattern and therefore is placed on the top left (where performance trend is positive and affect trend is negative). Producing this bottom scatter plot tells us that the relationship between affect and performance trend is negative. That is, people with a positive affect trend usually have a negative performance trend, people with a negative affect trend are more likely to have a positive performance trend, and people with no affect trend usually have no performance trend. 

\begin{center}

------------

Insert Figure \ref{trend_correlation} about here

------------

\end{center}

>> **Inference 3:** There are correlated trends.

The final trend inference is about identifying covariates or predictors of trend. Do helping behaviors predict the trend in depletion? Does the trend in unit climate covary with perceptions of leader quality? 

Notice the difference between this inference and inference three. Inference three asks how one trend is related to another, whereas this inference asks how raw values relate to or predict the trend in a different variable.

Figure \ref{trend_relation} (no figure yet). Paragraph about graph. **Not sure how to graph. Slope is a single term, like 0.5. It is being predicted by the raw values of X across time...** 

>> **Inference 4:** There are correlates/predictors of trend. 

## Trend Inference Table

The inference table below provides examples of each trend inference. Inference one is about the general direction or shape of a trajectory across time. Inference two is about variability in that shape across units. Inference three takes the trend in one variable and asks whether it tends to co-occur with trend in another. Inference four, finally, is about the relationship between trend in one variable and the raw values of one or more predictors.

```{r, echo = F}
library(xtable)
library(knitr)
library(kableExtra)
library(tidyverse)

inference_text2 <- c('There is positive/negative/curvilinear trend in a variable across time.?',
                    'There is variability in the trend of a variable across time. Trend differs across units.',
                    'There are correlated trends. There is a relationship between two trends.',
                    'There are correlates/predictors of trend.')
trend_table <- data.frame(
  Inference = c(1, 2, 3, 4),
  Examples = c(linebreak('Burnout decreases over time.\nPerformance increases over time.'),
                 linebreak('Affect trends differ across people (units).\nThere is variability in turnover trends across organizations.'),
                 linebreak('People with a positive health status trend have a positive happiness trend.\nPeople with a positive performance trend have a negative anxiety trend.'),
                 linebreak('Affect relates to the performance trend across time.\nHelping behaviors predict depletion trends.'))
)

trend_table %>%
  kable('latex', booktabs = T, escape = F, align = 'l') %>%
    column_spec(1, width = '5em') %>%
    column_spec(2, width = '30em') %>%
  row_spec(1:3, hline_after = T)


```


We want to close this section with a note on phrasing. The inferences we explored in this section have to do with correlating trends, where a statement like "affect and performance trends covary, such that people with a negative affect trend have a positive performance trend" is appropriate. There is a less precise phrase that is easy to fall into -- and we have seen it used in our literature. Sometimes, researchers will correlate trends and then state, "when affect decreases performance goes up." We urge researchers to avoid this second statement because it is not clear if it refers to a static relationship about trends or a dynamic statement about how trajectories move across time. That is, the phrase "when affect decreases performance goes up" could refer to correlated trends, but it could also mean something like, "when affect decreases performance immediately or subsequently goes up." This second statement is far different and it should not be used when an analysis only correlates trends or evokes predictors of trend. Again, we urge researchers to phrase their inferences as we have shown here.


```{r, echo = F, fig.cap = 'Trend across time\\label{trend}'}
set.seed(3)

ind <- c(1:20)
chris <- 4 + 0.7*ind + rnorm(20,0,0.5)
josh <- 4 - 0.7*ind  + rnorm(20, 0, 0.5)
chris[1] <- 4
josh[1] <- 4
#timothy <- c(4, 5, 15, 18, 21, 20, 19, 17, -6, -10, -12, -16, -14, 0, 1,2,3,4,5,9)
timothy <- 15*sin(ind/3)

fam <- data.frame(
  'Time' = c(ind, ind, ind),
  'Y' = c(chris, josh, timothy),
  'legend' = c(rep('Positive Trend', 20), rep('Negative Trend', 20), rep('Curvilinear Trend', 20))
)

hm <- data.frame(
  'Time' = c(ind),
  'Y' = c(timothy)
)

library(ggplot2)
ggplot(fam, aes(x = Time, y = Y, color = legend)) + 
  geom_point(alpha = 0.6) + 
  geom_abline(intercept = 4, slope = 0.7, color = 'blue', alpha = 0.5) +
  geom_abline(intercept = 4, slope = -0.7, color = 'red', alpha = 0.5) + 
  scale_y_continuous(labels = NULL) + 
  theme_classic() + 
  theme(legend.title = element_blank()) + 
  scale_color_manual(values = c('Positive Trend' = 'blue', 
                                'Negative Trend' = 'red',
                                'Curvilinear Trend' = 'green'),
                     breaks = c('Positive Trend', 
                                'Negative Trend',
                                'Curvilinear Trend')) + 
  geom_smooth(data = hm, aes(x = Time, y = Y), se = F, color = 'green', alpha = 0.6, size = 0.5, alpha = 0.2)


```

```{r, echo = F, fig.cap = 'Differences in trend variability across units\\label{trend_var}'}
set.seed(88)

per1 <- 4 + 0.6*ind + rnorm(20, 0, 0.5)
per2 <- 4 + 0.6*ind + rnorm(20, 0, 2)
per3 <- 4 + 0.6*ind + rnorm(20, 0, 0.5)

per4 <- 4 + 0.6*ind + rnorm(20, 0, 1)
per5 <- 4 - 0.9*ind + rnorm(20, 0, 1)
per6 <- 4 + 0.01*ind + rnorm(20,0,1)


yup <- data.frame(
  'Time' = c(ind, ind, ind, ind, ind, ind),
  'Y' = c(per1, per2, per3, per4, per5, per6),
  'Person' = c(rep('Person 1', 20),
               rep("Person 2", 20),
               rep('Person 3', 20),
               rep('Person 1', 20),
               rep('Person 2', 20),
               rep('Person 3', 20)),
  'type' = c(rep('No Trend Variability', 60),
             rep('Trend Variability', 60))
)


ggplot(yup, aes(x = Time, y = Y, color = Person)) + 
  geom_point(alpha = 0.6) + 
  geom_smooth(method = 'lm', se = F, aes(color = Person), size = 0.5, alpha = 0.6) + 
  facet_wrap(~type) + 
  theme_classic() + 
  theme(legend.title = element_blank()) + 
  scale_y_continuous(labels = NULL) + 
  scale_color_manual(values = c('Person 1' = 'red',
                                'Person 2' = 'green',
                                'Person 3' = 'blue')
                     )



```


```{r, echo = F, fig.cap = 'Correlating slopes, or relating the affect to performance trend\\label{trend_correlation}', fig.height = 10}

set.seed(33)
tre_affect <- 6 + 0.6*ind + rnorm(20,0,1)
tre_perf <- 2 - 0.6*ind + rnorm(20,0,1)

affect2 <- 6- 0.6*ind + rnorm(20,0,1)
perf2 <- 2 + 0.6*ind + rnorm(20,0,1)

affect3 <- 5 + 0.02*ind + rnorm(20,0,1)
perf3 <- 5 + 0.01*ind + rnorm(20,0,1)

trend_rel <- data.frame(
  'Time' = c(ind, ind,
             ind, ind,
             ind, ind),
  'Value' = c(tre_affect, tre_perf,
          affect2, perf2,
          affect3, perf3),
  'call' = c(rep('Affect', 20),
              rep('Performance', 20),
             rep('Affect', 20),
             rep('Performance', 20),
             rep('Affect', 20),
             rep('Performance', 20)),
  'Person' = c(rep('Person 1', 40),
               rep('Person 2', 40),
               rep('Person 3', 40))
)

trend_lines <- data.frame(
  'Person' = c('Person 1',
          'Person 1',
          'Person 2',
          'Person 2',
          'Person 3',
          'Person 3'),
  'call' = c('Affect',
             'Performance',
             'Affect', 
             'Performance',
             'Affect',
             'Performance'),
  'intercept' = c(6, 2,
                  6, 2, 
                  5, 5),
  'slope' = c(0.6, -0.6,
              -0.6, 0.6,
              0.02, 0.01)
)

tp1 <- ggplot(trend_rel, aes(x = Time, y = Value, group = call)) + 
  geom_line(aes(linetype = call)) + 
  theme_classic() + 
  scale_y_continuous(labels = NULL) + 
  theme(legend.title = element_blank()) + 
  facet_grid(~Person, scales = 'free', space = 'free') + 
  geom_abline(aes(intercept = intercept, slope = slope, color = Person), data = trend_lines, alpha = 0.9, show.legend = F) + 
  scale_color_manual(values = c('Person 1' = 'red',
                                'Person 2' = 'green',
                                'Person 3' = 'blue'))
  



trend_trend <- data.frame(
  'Affect' = c(-0.8, 0, 0.8),
  'Performance' = c(0.8, 0, -0.8),
  'label' = c('green', 'blue', 'red')
)

tp2 <- ggplot(trend_trend, aes(x = Affect, y = Performance, color = label)) + 
  geom_point(size = 6, alpha = 0.6) + 
  theme_classic() + 
  scale_y_continuous(breaks = c(-1, 0, 1)) + 
  scale_x_continuous(breaks = c(-1, 0, 1)) + 
  ylab("Trend in Performance") + 
  xlab("Trend in Affect") + 
  theme(legend.title = element_blank()) + 
  scale_color_manual(values = c('red' = 'red', 
                                'blue' = 'blue',
                                'green' = 'green'),
                     labels = c('Person 3',
                                'Person 2',
                                'Person 1'))
                     #breaks = c('Person 1',
                          #      'Person 2',
                           #     'Person 3'))

library(gridExtra)

grid.arrange(tp1, tp2, ncol = 1)

```



## Models

Trends are called slope estimates in the statistical modeling literature. They are also referred to as growth. Mean estimates of slopes, or trends, or growth will tell you about trend, whereas the variance estimates will tell you about variability across units. Growth curves in SEM or HLM. Bivariate growth curves. 



